---
title: "NFL_Analysis"
output: 
  html_document:
    toc: true
    theme: cosmo
---

<!-- I have no idea what below does -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We're going to do some basic analysis on NFL teams from 2002-2020. This consists
of the 16-game-per-season era.

Let's import some important libraries.
```{r imports}
library(tidyverse)
library(ggplot2)
```

For me, the dataset is already in the directory. You can also download it straight
off the link
```{r get data}
df <- read_csv('standings.csv')
# data <- read_csv('http://www.habitatring.com/standings.csv') # if you want the data from online
```

First, we will conduct some exploration and try to understand the dataset first.
Then, we will explore the relationship of points. We'll explore points scored,
points allowed, and point differential versus wins. Obviously, for individual
games, measurements like point differential are highly correlated with wins, but
over the average course of a season we might see different results.

Note: After the analysis, it makes more sense to plot pct instead (I think
it takes into account draws as well). Overall the graphs should still look the same.

Finally, I would like to conduct a search for outliers of these statistics. For example,
which team is the most 'clutch', that is, has the highest win rate with a low point
differential. We could see the clutchest team in a season or the clutchest overall
NFL team throughout the 16-game era. Alternatively, we could also explore the least
clutch team, the "chokers". 

## Data analysis

Let's first get a look of what's in our data!
```{r first look}
head(df)
names(df)
summary(df)
```

There are honestly a lot of columns. For now, let's filter our data. What columns
should we keep? season, team, wins, losses, ties, pct, scored, allowed, and net all make sense.
Let's also keep seed, it might come in handy later.

```{r filter}
df <- select(df, c(season, team, wins, losses, ties, pct, scored, allowed, net, seed))
head(df)
```

There are 32 teams in the NFL, and 19 years in this dataset. Let's create an aggregated
dataset where we combine all of a teams stats.
```{r combined dataset}
combined_df <- df %>%
  group_by(team) %>%
  summarise(wins = sum(wins), losses = sum(losses), ties = sum(ties),
            pct = mean(pct), scored = sum(scored), allowed = sum(allowed),
            net = sum(net), seed = mean(seed))
print(combined_df)
```
Note: I chose to sum up all the data, except in pct and seed, where I thought
it would make more sense to have them be averages.

Alright, now it's time to do some very basic plotting. Let's just look at scored
vs pct in the original data.
```{r scored vs wins original}
ggplot(data=df, aes(x = scored, y = pct)) + geom_point()
```

Very interesting. For the most part, of course we see a direct positive correlation. 
The more you score, the more you win, right? But what about the line on the left, which
essentially has a bunch of super low scoring team. My hypothesis is that some of
our data is missing, so let's go ahead and split by year.

```{r scored vs wins original by year}
ggplot(data=df, aes(x = scored, y = pct, color=factor(season))) + geom_point()
```

Okay, the colors are confusing, but I think we can clearly tell that there's something
wrong with the data from 2020 or so. Let's plot the filtered data from 2016 onwards.

```{r data from 2018+}
df_2016_plus <- subset(df, df$season >= 2016)
ggplot(data=df_2016_plus, aes(x = scored, y = pct, color=factor(season))) + geom_point()
```

Now it's very clear that we only have problems with 2020. Let's inspect the data.

```{r data from 2020}
df_2020 <- subset(df, df$season == 2020)
head(df_2020)
summary(df_2020)
```

It's clear that our data is a bit outdated and doesn't have everything in the 2020 season.
Let's update our dataset to only include analysis from 2002-2019 for the graphs.
Let's update the combined dataset as well. (as means may be lopsided by overweighting 2020).

```{r cleaning data}
df <- subset(df, df$season < 2020)
combined_df <- df %>%
  group_by(team) %>%
  summarise(wins = sum(wins), losses = sum(losses), ties = sum(ties),
            pct = mean(pct), scored = sum(scored), allowed = sum(allowed),
            net = sum(net), seed = mean(seed))
head(df) #to check
head(combined_df) # to check
```

Now let's go ahead and plot scored, allowed, and net vs pct for our data.

```{r general plots}
ggplot(data=df, aes(x=scored, y=pct)) + geom_point(color=factor(df$season)) + geom_smooth()
ggplot(data=df, aes(x=allowed, y=pct)) + geom_point(color=factor(df$season)) + geom_smooth()
ggplot(data=df, aes(x=net, y=pct)) + geom_point(color=factor(df$season)) + geom_smooth()
```

Honestly, just as expected. There is a strong correlation between scored points,
allowed points, and an extremely strong correlation for net points. Let's calculate
the correlation exactly.

```{r correlations}
print(cor(x=df$scored, y=df$pct))
print(cor(x=df$allowed, y=df$pct))
print(cor(x=df$net, y=df$pct))
```

Interestingly enough, defense is not as correlated.
There's too many dots on these graphs, and labeling would be a visual nightmare.
Let's switch over to the combined datasets. Let's look at the generic plots before
answering/asking any questions.

<!-- It would be really nice to make all these points different colors, but idk how. -->
```{r combined general plots}
ggplot(data=combined_df, aes(x=scored, y=pct)) + geom_point() + geom_text(aes(label=combined_df$team))
ggplot(data=combined_df, aes(x=allowed, y=pct)) + geom_point() + geom_text(aes(label=combined_df$team))
ggplot(data=combined_df, aes(x=net, y=pct)) + geom_point() + geom_text(aes(label=combined_df$team))
```

The data doesn't look that linear/correlated. It seems to be a problem with LAC and LA. Maybe they are new teams?
TODO: blah blah NE is too good, CLE sucks
TODO: correlation, correlation graph
TODO (?): clutch factor

## Conclusion

TODO

## Session Info
```{r session info}
sessionInfo()
```

